Supplementary Materials

### 2023-07-26 analysis/supplementary-materials/supp-mat.qmd
prepare supp mat for submission

---
title: "Supplementary Materials"
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    message: false
format: 
  html:
    toc: true
    code-fold: true
  pdf:
    toc: true
bibliography: "../paper/references.bib"
---

@@

@@

@@

The sample consists of `r nrow(demography)` individuals, most of which are
middle adult male individuals (@fig-age-distribution). Middle adult males were
preferentially targeted and the sample age and sex distribution is therefore not
representative of the population. This was done to limit potential confounding
factors, and because pipe notches, which served as a positive control for tobacco,
are predominantly seen in male individuals at the site.

@@

Missing data per dataset

```{r}
#| label: fig-missing-caries
caries %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums()

caries %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id) %>% 
  ggplot(aes(x = name, y = id, fill = value)) +
    geom_tile()
```

Missing data across all caries, periodontitis, and periapical datasets.

```{r}
#| label: fig-missing-dental
#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
dental_long %>%
  mutate(across(c(caries, periodont, periap), is.na)) %>%
  select(c(caries, periodont, periap, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

# successfully replicated samples only
uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  mutate(compound = str_remove(compound, "_calc")) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) #%>%
  # filter(presence == 0 | presence == 2) %>%
  # group_by(compound) %>%
  # mutate(presence = if_else(presence == 2, 1, 0)) # convert replications to presence/absence

uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
  #right_join(select(filter(metadata, replicated == TRUE)), by = c("id", "sample"))
  #mutate(compound = str_remove(compound, "_calc")) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) %>%
  filter(presence == 0 | presence == 2) %>% # remove compounds only detected in one batch
  group_by(compound) %>%
  mutate(presence = if_else(presence == 0, 0, 1)) %>%  # convert replications to presence/absence
  ungroup()

@@

```{r}
#| label: tbl-uhplc-batch-1
#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch1") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  select(!c(sample, conc, extraction, batch, presence)) %>%
  pivot_wider(names_from = "compound", values_from = "quant") %>% 
  knitr::kable()
```

```{r}
#| label: tbl-uhplc-batch-2
#| tbl-cap: "Results from the UHPLC analysis second batch. Quantity of compound in calculus after third wash, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch2") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  pivot_wider(names_from = "compound", values_from = "quant") %>%
  select(!c(extraction, batch))
```

@@

```{r}
#| label: tbl-uhplc-replicated
uhplc_calculus_replicated %>%
  filter(sum(presence) != 0) %>% # remove compounds absent in all individuals in sample
  pivot_wider(names_from = "compound", values_from = "presence")
```

<!-- line plot with concentration across washes and to the final calculus sample -->

@@

```{r auth-plot-batch1}
#| label: fig-auth-plot-batch1

uhplc_data_long %>%
  filter(
    batch == "batch1",
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

```{r}
#| label: fig-auth-plot-batch2
uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw()
```

@@

The positive correlation between the weight of the calculus
sample and recovered quantities of the compounds suggests sample weight may affect
the ability to detect compounds; although, we were able to detect
compounds in samples as small as 2 mg [@fig-quant-weight-1 and @fig-quant-weight-2].

```{r}
#| label: fig-quant-weight-1
#| #| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 1."
uhplc_data_comb %>%

  select(sample, contains("batch1")) %>% 
  select(sample, batch1_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch1_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  filter(conc > 0) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  ggplot(aes(x = batch1_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "", y = "Quantity (ng)")
```

```{r}
#| label: fig-quant-weight-2
#| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 2."
uhplc_data_comb %>%
  select(sample, contains("batch2")) %>% 
  select(sample, batch2_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch2_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  #remove_missing() %>% 
  filter(conc > 0) %>% 
  ggplot(aes(x = batch2_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    labs(y = "Quantity (ng)", y = "", col = "Sample number")
```

@@

# awkward wrangling required because the information on which individuals were included in replication was lost when combining batches 1 and 2
#n_batch1 <- length(metadata$replicated)
#n_batch2 <- sum(metadata$replicated) # sum boolean vector to get length of TRUE values
uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  #mutate(detect = replace_na(conc, 0)) %>% 
  #mutate(detect = if_else(detect > 0, 1, 0)) %>%
  #group_by(compound, batch) %>% 
  # summarise(
  #   n_detect = sum(detect)
  #           ) %>% 
  # mutate(n = case_when(batch == 1 ~ n_batch1,
  #                      batch == 2 ~ n_batch2)) %>% 
  # mutate(prop = n_detect / n) %>%

ggplot(aes(x = compound, fill = compound)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

@@

uhplc_calculus_long %>%
  filter(id %in% filter(metadata, replicated == T)$id,
         quant > 0) %>%
  #remove_missing() %>%
  ggplot(aes(x = compound, fill = compound)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

@@

```{r}
#| label: fig-detection-preservation
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    batch == "batch1",
    quant > 0,
    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()
```

```{r}
#| label: fig-detection-preservation2
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    quant > 0,
    batch == "batch2",
    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

@@

```{r}
#| label: fig-preservation-detection
uhplc_calculus_long %>% 
  ungroup() %>% 
  distinct(id, batch, .keep_all = TRUE) %>%
  remove_missing(vars = c("preservation", "weight")) %>% 
  ggplot(aes(x = preservation)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw()
```

@@

```{r}
#| label: fig-corr-notches
# correlation between nicotine and cotinine conc, and pipe notches
uhplc_calculus_long %>%
  filter(compound %in% c("caffeine", "cotinine", "nicotine", "salicyl", "theophyl")) %>% 
  ggplot(aes(x = pipe_notch, y = conc)) +
    geom_point() +
    facet_wrap(~ compound + batch, scales = "free_y")
```

@@

tobacco_comb <- tobacco %>% 
  group_by(batch, sample, id, .add = TRUE) %>%
  summarise(
    detection = sum(detection), 
    #sex, age, completeness, preservation, pipe_notch, 
    .groups = "keep" # combine compounds: 0 = none detected; 1 = 1 detected; 2 = both detected
  ) %>%
  ungroup() %>%
  left_join(select(demography, id, pipe_notch), by = "id") %>%
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection > 0 & pipe_notch == "Y" ~ TRUE,
      detection == 0 & pipe_notch == "N" ~ TRUE,
      detection > 0 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
      TRUE ~ FALSE
      )
    )

@@

# accuracy in replicated samples
uhplc_accuracy <- uhplc_calculus_replicated %>%
  left_join(select(demography, id, pipe_notch), by = "id") %>% 
  filter(compound == "nicotine" | compound == "cotinine") %>%
  group_by(id) %>% 
  arrange(desc(presence), .by_group = T) %>%
  distinct(id, .keep_all = T) %>% 
  #slice(which(str_detect(compound, "nicotine|cotinine"))) #%>%
  # slice(which(compound == "nicotine" | compound == "cotinine")) %>%
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(presence == 1 & pipe_notch == "Y" ~ TRUE,
                             presence == 0 & pipe_notch == "N" ~ TRUE, # can we be sure if this is correct?
                             presence == 1 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
                             TRUE ~ FALSE))

@@

```{r}
#| label: fig-notch-ratio
#| fig-cap: "Relationship between ratio and number of pipe notches."

tobacco_ratio %>%
  ggplot(aes(y = ratio, x = pipe_notch)) +
    #geom_boxplot()
    geom_point()

  #lm(ratio ~ age + sex + preservation + pipe_notch, tobacco_ratio)
```

@@

No apparent correlation between the number of pipe notches
and the concentration of nicotine or cotinine [@fig-corr-notches].

@@

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
out of
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.

Nicotine and cotinine detection successfully identified
`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y"))`
(`r scales::percent(nrow(filter(tobacco_comb, pipe_notch == "Y", batch == 1, detection > 0)) / nrow(filter(tobacco_comb, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`)
individuals with pipe notches in the first analysis.

Nicotine detection successfully identified
`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(nicotine, batch == 1, pipe_notch == "Y"))`
individuals with pipe notches
(`r scales::percent(nrow(filter(nicotine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(nicotine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).

Cotinine detection successfully identified
`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0))`
smokers out of
`r nrow(filter(cotinine, batch == 1, pipe_notch == "Y"))`
individuals with pipe notches
(`r scales::percent(nrow(filter(cotinine, batch == 1, pipe_notch == "Y", detection > 0)) / nrow(filter(cotinine, batch == 1, pipe_notch == "Y")), accuracy = 0.1)`).

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
out of
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
smokers
(`r scales::percent(mean(filter(uhplc_accuracy, pipe_notch == "Y")$correct, na.rm = T))`).
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.

@@

```{r}
#| label: fig-dental-plots
#| layout-ncol: 2
#| fig-cap: 
#|   - "Antemortem tooth loss in sample."
#|   - "Periodontal scores in the sample."
dental_long %>%
  mutate(status = case_when(status == "dna" ~ "p")) %>% 
  dental_plot(fill = status)

dental_long %>%
  filter(!is.na(periodont)) %>% 
  dental_plot(fill = periodont)
```

@@

Ratio of antemortem lost teeth per present teeth at the site.

```{r}
#| label: fig-aml-id
#| fig-cap: "AMTL ratio per individual."
dental_long %>% 
  group_by(id) %>% 
  amtl_ratio(.status = status, .add = T)
```

```{r}
#| label: fig-aml-class
#| fig-cap: "AMTL ratio per tooth class."
dental_long %>% 
  group_by(class) %>% 
  amtl_ratio(.status = status, .add = T)
```

```{r}
#| label: fig-aml-type
#| fig-cap: "AMTL ratio per tooth type."
dental_long %>% 
  group_by(type) %>% 
  amtl_ratio(.status = status, .add = T)
```

@@

<!--add a caries correction?-->
Caries were scored as the location on each individual tooth. Multiple locations
on a single tooth were separated with `;`. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces with an unknown origin were recorded as 'crown'.

@@

```{r}
surface <- c("mes", "dis", "occ", "buc", "lin", "root", "crown") # here occ = incisal

# caries_count <- dental_long %>% 
#   select(!c(periodont, periap, calculus)) %>% 
#   #filter(status == "p") %>%
#   separate_rows(caries, sep = ";") %>% # one lesion per row
#   mutate(caries_count = if_else(
#     caries == "none", 0L, 1L
#     )
#   ) %>%
#   group_by(id, tooth) %>%
#   summarise(
#     caries_count = sum(caries_count, na.rm = F)
#     ) %>%
#   tooth_position()
# 
# site_wide_caries <- caries_count %>%
#   caries_ratio(.caries = caries_count)
```

>>>>

---
title: "Supplementary Materials"
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    message: false
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: true
bibliography: "../paper/references.bib"
---

@@

These supplementary figures and tables are a variable hodgepodge of things that
didn't fit in the main manuscript, and additional things I thought might be
useful. The best way to explore/verify the results and interpretations is to
download all the data and code (<https://doi.org/10.5281/zenodo.7649824>)
and just play around with it yourself. Enjoy!

@@

Calculus samples are a combination of leftovers from a previous aDNA study and
newly sampled individuals. In some cases individuals from the previous study were
sampled again (if not enough calculus was left over from the previous study)
(@tbl-sampling).

@@

The sample consists of `r nrow(demography)` individuals, most of which are
middle adult male individuals (@fig-age-distribution). Middle adult males were
preferentially targeted due to larger calculus deposits (observation) and the
sample age and sex distribution is therefore not
representative of the population. This was also done to limit potential confounding
factors, and because pipe notches, which served as a positive control for tobacco,
are predominantly seen in male individuals at the site.

@@

An overview of the missing teeth can be found in @fig-missing-teeth. Missing
scores per tooth can be found in @tbl-missing-scores and @fig-missing-dental.

```{r}
#| label: fig-missing-teeth
#| fig-cap: "Heatmap of missing teeth per individual in the sample. 1 = present, 2 = missing."
dental_inv_long %>% 
  mutate(status = if_else(status == "p", 1, 0)) %>% 
  ggplot(aes(x = tooth, y = id, fill = status)) +
    geom_tile()
```

```{r}
#| label: missing-values-teeth
caries_missing <- caries %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums() %>% 
  as_tibble_row()

periodont_missing <- periodont %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums() %>% 
  as_tibble_row()

periap_missing <- periap %>%
  mutate(across(!id, ~ is.na(.x))) %>%
  select(!id) %>%
  colSums() %>% 
  as_tibble_row()

missing_tbl <- tibble(caries_missing) %>% 
  add_case(periodont_missing) %>% 
  add_case(periap_missing) %>% 
  mutate(score = c("caries", "periodontitis", "periapical"), .before = 1)
```

```{r}
#| label: tbl-missing-scores
#| tbl-cap: "Table of missing scores by tooth."
knitr::kable(missing_tbl)
```

```{r}
#| label: fig-missing-dental
#| fig-cap: "Plots"
#| layout-ncol: 2
#| fig-subcap: 
#|   - "Caries"
#|   - "Periodontitis"
#|   - "Periapical lesions"
#|   - "Combined caries, periodontitis, periapical."
caries %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id, names_to = "tooth") %>% 
  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
  ggplot(aes(x = tooth, y = id, fill = value)) +
    geom_tile()

periodont %>%
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id, names_to = "tooth") %>% 
  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
  ggplot(aes(x = tooth, y = id, fill = value)) +
    geom_tile()

periap %>% 
  mutate(across(!id, ~ is.na(.x))) %>% 
  pivot_longer(-id, names_to = "tooth") %>% 
  mutate(value = if_else(value == TRUE, "missing", "present")) %>% 
  ggplot(aes(x = tooth, y = id, fill = value)) +
    geom_tile()

dental_long %>%
  mutate(across(c(caries, periodont, periap), is.na)) %>%
  select(c(caries, periodont, periap, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

# successfully replicated samples only
uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  mutate(compound = str_remove(compound, "_calc")) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) #%>%

uhplc_calculus_replicated <- uhplc_calculus_bin %>%
  filter(id %in% filter(metadata, replicated == TRUE)$id) %>%
  group_by(id, sample, compound) %>% # combine batches
  summarise(presence = sum(presence)) %>%
  filter(presence == 0 | presence == 2) %>% # remove compounds only detected in one batch
  group_by(compound) %>%
  mutate(presence = if_else(presence == 0, 0, 1)) %>%  # convert replications to presence/absence
  ungroup()

@@

```{r}
#| label: tbl-uhplc-batch-1
#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch1") %>%
  select(!c(sample, conc, extraction, batch, presence)) %>%
  pivot_wider(names_from = "compound", values_from = "quant") %>% 
  knitr::kable()
```

```{r}
#| label: tbl-uhplc-batch-2
#| tbl-cap: "Results from the UHPLC analysis second batch. Quantity of compound in calculus after third wash, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch2") %>%
  pivot_wider(names_from = "compound", values_from = "quant") %>%
  select(!c(extraction, batch)) %>% 
  knitr::kable()
```

@@

@@

```{r}
#| label: fig-auth-plot-batch1
#| fig-cap: "Plot of extracted quantities of each compound across the three washes and calculus extraction in batch 1. Each line represents an individual."
uhplc_data_long %>%
  filter(
    batch == "batch1",
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y", ncol = 3) +
    theme_bw()
```

```{r}
#| label: fig-auth-plot-batch2
#| fig-cap: "Plot of extracted quantities of each compound across the three washes and calculus extraction in batch 2. Each line represents an individual."
uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    facet_wrap(~ compound, scales = "free_y", ncol = 2) +
    theme_bw()
```

@@

The positive correlation between the weight of the calculus
sample and recovered quantities of the compounds suggests sample weight may affect
the ability to detect compounds; although, we were able to detect
compounds in samples as small as 2 mg (@fig-quant-weight-1 and @fig-quant-weight-2).

```{r}
#| label: fig-quant-weight-1
#| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 1."
uhplc_data_comb %>%

  select(sample, contains("batch1")) %>% 
  select(sample, batch1_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch1_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  filter(conc > 0) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  ggplot(aes(x = batch1_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = "Calculus weight (mg)", y = "Quantity (ng)")
```

```{r}
#| label: fig-quant-weight-2
#| fig-cap: "Quantity of a compound (ng) found in a sample plotted against the weight of the calculus sample. Results from batch 2."
uhplc_data_comb %>%
  select(sample, contains("batch2")) %>% 
  select(sample, batch2_weight, contains("calc")) %>% 
  pivot_longer(
    -c(sample, batch2_weight),
    names_to = c("compound", "batch"),
    names_pattern = "(.*)_(.*)",
    values_to = c("conc")
  ) %>% 
  mutate(compound = str_remove(compound, "_calc")) %>%
  #remove_missing() %>% 
  filter(conc > 0) %>% 
  ggplot(aes(x = batch2_weight, y = conc, col = as.factor(sample))) +
    geom_point() +
    facet_wrap(~ compound, scales = "free_y") +
    theme_bw() +
    labs(y = "Quantity (ng)", x = "Calculus weight (mg)", col = "Sample number")
```

@@

uhplc_calculus_long %>%
  filter(quant > 0) %>% 
ggplot(aes(x = compound, fill = compound)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

@@

uhplc_calculus_long %>%
  filter(id %in% filter(metadata, replicated == T)$id,
         quant > 0) %>%
  ggplot(aes(x = compound, fill = compound)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "none"
      )
```

@@

```{r}
#| label: fig-detection-preservation
#| fig-cap: "Plot of relationship between the absolute quantity of a detected compound (ng) and the overall skeletal preservation of the individuals in which the compound was detected. Showing results for batch 1."
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    batch == "batch1",
    quant > 0,
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()
```

```{r}
#| label: fig-detection-preservation2
#| fig-cap: "Plot of relationship between the absolute quantity of a detected compound (ng) and the overall skeletal preservation of the individuals in which the compound was detected. Showing results for batch 2."
uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    quant > 0,
    batch == "batch2",
    #!compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva"), # remove compounds not detected
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

@@

```{r}
#| label: fig-preservation-detection
#| fig-cap: "Plot of the number of skeletons in each state of preservation separated by batch."
uhplc_calculus_long %>% 
  ungroup() %>% 
  distinct(id, batch, .keep_all = TRUE) %>%
  remove_missing(vars = c("preservation", "weight")) %>% 
  ggplot(aes(x = preservation)) +
    geom_bar() +
    facet_wrap(~ batch) +
    theme_bw()
```

@@

```{r}
#| label: fig-corr-notches
#| fig-cap: "Plot of relationships between the total number of pipe notches in an individual and the concentration of detected compounds. The only relevant comparisons in this case are nicotine and cotinine. The others are just included because I couldn't be bothered filtering them out."
# correlation between nicotine and cotinine conc, and pipe notches
uhplc_calculus_long %>%
  filter(compound %in% c("caffeine", "cotinine", "nicotine", "salicyl", "theophyl")) %>% 
  ggplot(aes(x = pipe_notch, y = conc)) +
    geom_point() +
    facet_wrap(~ compound + batch, scales = "free_y") +
    theme_bw() +
    labs(x= "number of pipe notches", y = "detected compound concentration (ng/mg)")
```

@@

tobacco_comb <- tobacco %>% 
  group_by(batch, sample, id, .add = TRUE) %>%
  summarise(
    detection = sum(detection), 
    .groups = "keep" # combine compounds: 0 = none detected; 1 = 1 detected; 2 = both detected
  ) %>%
  ungroup() %>%
  left_join(select(demography, id, pipe_notch), by = "id") %>%
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(
      detection > 0 & pipe_notch == "Y" ~ TRUE,
      detection == 0 & pipe_notch == "N" ~ TRUE,
      detection > 0 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
      TRUE ~ FALSE
      )
    )

@@

# accuracy in replicated samples
uhplc_accuracy_batch2 <- uhplc_calculus_long %>%
  filter(compound == "nicotine" | compound == "cotinine",
         batch == "batch2") %>%
  group_by(id) %>% 
  arrange(desc(presence), .by_group = T) %>%
  distinct(id, .keep_all = T) %>% 
  mutate(
    pipe_notch = if_else(pipe_notch > 0, "Y", "N"),
    correct = case_when(presence == 1 & pipe_notch == "Y" ~ TRUE,
                             presence == 0 & pipe_notch == "N" ~ TRUE, # can we be sure if this is correct?
                             presence == 1 & pipe_notch == "N" ~ NA, # can't be sure if this is true or not
                             TRUE ~ FALSE))

@@

@@

No apparent correlation between the number of pipe notches
and the concentration of nicotine or cotinine (@fig-corr-notches).

@@

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y", presence > 0))`
out of
`r nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy_batch2, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(uhplc_accuracy_batch2$correct, na.rm = T), accuracy = 0.1)`.

@@

An overview of available teeth can be seen in @fig-dental-inv.

```{r}
#| label: fig-dental-inv
#| fig-cap: "Overview of the dental inventory of the sample. Teeth removed for DNA analysis considered 'present'."
dental_long %>%
  mutate(status = case_when(status == "dna" ~ "p", TRUE ~ status)) %>% 
  dental_plot(fill = status)
```

@@

Ratios of antemortem lost teeth per present teeth at the site. Calculated per
individual (@tbl-aml-id), tooth class (@tbl-aml-class), and tooth type (@tbl-aml-type)

```{r}
#| label: tbl-aml-id
#| tbl-cap: "AMTL ratio per individual."
dental_long %>% 
  group_by(id) %>% 
  amtl_ratio(.status = status, .add = T) %>% 
  knitr::kable()
```

```{r}
#| label: tbl-aml-class
#| tbl-cap: "AMTL ratio per tooth class."
dental_long %>% 
  group_by(class) %>% 
  amtl_ratio(.status = status, .add = T) %>% 
  knitr::kable()
```

```{r}
#| label: tbl-aml-type
#| tbl-cap: "AMTL ratio per tooth type."
dental_long %>% 
  group_by(type) %>% 
  amtl_ratio(.status = status, .add = T) %>% 
  knitr::kable()
```

@@

Caries were scored as the location on each individual tooth. Multiple locations
on a single tooth were separated with `;`. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces with an unknown origin were recorded as 'crown'.

@@

In the `r caries_ratio_site$n_teeth` that were examined, `r caries_ratio_site$count`
teeth had caries
(`r scales::percent(caries_ratio_site$ratio, accuracy = 0.1)`).
This frequency has very little meaning, and was further broken down into a ratio
for each individual and each tooth class (@tbl-caries-id-class and @fig-caries-class).
As expected, the molars have a higher frequency of caries than the other teeth.

```{r}
#| label: tbl-caries-id-class
#| tbl-cap: "Table of caries ratios per individual per tooth class."
caries_count %>%
  caries_ratio(.caries = count, id, class) %>% 
  knitr::kable()
```
